---
title: "Exploring the McGill XevaSet with Xeva"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
output:
  html_document:
    toc: true
    toc_depth: 3
    mathjax: null
---

This document introduces the curated McGill breast cancer PDX `XevaSet` using the
`Xeva` R package and the exported pipeline tables.

## Setup

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.width = 8,
  fig.height = 5
)

suppressPackageStartupMessages({
  library(Biobase)
  library(data.table)
  library(dplyr)
  library(ggplot2)
  library(knitr)
  library(patchwork)
  library(scales)
  library(tidyr)
  library(Xeva)
})

theme_set(theme_minimal(base_size = 12))

data_path <- function(...) file.path("data", ...)
```

## Load The Dataset

The curated object is stored as an `RDS` file in `data/results/`.

```{r load-data}
xeva_path <- data_path("results", "Xeva_McGill.rds")

mcgill <- readRDS(xeva_path)
mcgill
```

The `XevaSet` bundles model metadata, drug metadata, tumor growth
measurements, response summaries, and molecular profiles into one object.

```{r dataset-overview}
model_df <- modelInfo(mcgill) |>
  as_tibble()
treatment_df <- mcgill@drug |>
  as_tibble()
batch_ids <- batchInfo(mcgill)
model_metrics <- mcgill@sensitivity$model |>
  as.data.frame(stringsAsFactors = FALSE)
batch_metrics <- mcgill@sensitivity$batch |>
  as.data.frame(stringsAsFactors = FALSE)
manifest <- fread(data_path("results", "Xeva_McGill_csv", "file_manifest.csv"))
mod_map <- fread(data_path("results", "Xeva_McGill_csv", "mod_to_biobase_map.csv"))
example_experiment <- getExperiment(mcgill, model.id = model_df$model.id[[1]])

overview_tbl <- tibble(
  quantity = c(
    "Model records",
    "Patients",
    "Treatment-control batches",
    "Treatments including control",
    "Tumor measurements",
    "Molecular profile types"
  ),
  value = c(
    nrow(model_df),
    n_distinct(model_df$patient.id),
    length(batch_ids),
    n_distinct(treatment_df$drug.id),
    manifest$rows[manifest$component == "experiment_measurements"],
    length(mcgill@molecularProfiles)
  )
)

kable(overview_tbl)
```

## What Is Included?

```{r object-contents}
contents_tbl <- tibble(
  component = c(
    "Model metadata",
    "Treatment metadata",
    "Individual growth curves",
    "Model-level response summaries",
    "Batch-level response summaries",
    "Molecular profiles",
    "CSV exports"
  ),
  rows_or_items = c(
    nrow(model_df),
    nrow(treatment_df),
    length(mcgill@experiment),
    nrow(model_metrics),
    nrow(batch_metrics),
    length(mcgill@molecularProfiles),
    nrow(manifest)
  ),
  fields_or_summary = c(
    paste(colnames(model_df), collapse = ", "),
    paste(colnames(treatment_df), collapse = ", "),
    paste(colnames(example_experiment), collapse = ", "),
    paste(colnames(model_metrics), collapse = ", "),
    paste(colnames(batch_metrics), collapse = ", "),
    paste(names(mcgill@molecularProfiles), collapse = ", "),
    paste(manifest$relative_path, collapse = ", ")
  )
)

kable(contents_tbl)
```

```{r growth-curve-table-example}
kable(head(example_experiment, 8))
```

## Workbook Parsing

The treatment-response workbook is parsed into one model-level row per
mouse-treatment arm and one measurement row per time point.

```{r workbook-summary}
block_summary <- fread(data_path(
  "procdata", "metadata", "tumor_workbook_blocks.csv"
))

workbook_summary <- block_summary |>
  summarise(
    patients = n_distinct(patient.id),
    model_records = n(),
    control_records = sum(drug == "untreated"),
    treated_records = sum(drug != "untreated"),
    min_time = min(min_time, na.rm = TRUE),
    max_time = max(max_time, na.rm = TRUE)
  )

kable(workbook_summary)
```

```{r workbook-treatment-summary}
workbook_treatment_summary <- block_summary |>
  count(drug, name = "model_records") |>
  arrange(desc(model_records), drug)

kable(workbook_treatment_summary)
```

```{r treatment-barplot, fig.height=4}
ggplot(
  workbook_treatment_summary,
  aes(x = reorder(drug, model_records), y = model_records)
) +
  geom_col(fill = "#3b82b8") +
  coord_flip() +
  labs(
    title = "Model records by treatment",
    x = "Treatment",
    y = "Model records"
  )
```

## Model Metadata

Each row in the model metadata represents one mouse-level model in a treatment
context. Controls use `untreated` as the normalized drug label.

```{r model-metadata-head}
kable(head(model_df, 8))
```

Key metadata columns include:

* `model.id`: unique mouse-treatment model identifier.
* `patient.id`: patient-level `GCRC` identifier shared across related arms.
* `drug`: normalized treatment name.
* `source_treatment_name`: treatment label from the source workbook.
* `source.dose` and `source.schedule`: workbook dose and schedule where
  provided.
* `RNAseq.PDX` and `WGS.PDX`: molecular sample identifiers from the crosswalk.

```{r clinical-summary}
clinical_columns <- c("ER", "PR", "HER2", "AIMS")

clinical_summary <- model_df |>
  distinct(patient.id, across(all_of(clinical_columns))) |>
  pivot_longer(
    cols = all_of(clinical_columns),
    names_to = "field",
    values_to = "value"
  ) |>
  mutate(value = if_else(is.na(value), "Not available", value)) |>
  count(field, value, name = "patients") |>
  arrange(field, desc(patients), value)

kable(clinical_summary)
```

## Treatment Metadata

Treatment metadata preserves the source treatment label and includes compact
compound annotation fields where available.

```{r treatment-metadata}
kable(
  treatment_df |>
    select(
      drug.id,
      standard.name,
      source_treatment_name,
      treatment.type,
      source.patient.count,
      source.model.count,
      annotationdb.match,
      compound.name,
      pubchem.cid,
      molecular.formula,
      smiles,
      inchikey
    )
)
```

```{r unmatched-treatment-report}
unmatched_treatments <- fread(data_path("results", "unmatched_treatments.csv"))

kable(tibble(
  report = "Unmatched treatments",
  rows = nrow(unmatched_treatments)
))
```

```{r unmatched-treatment-table}
if (nrow(unmatched_treatments) > 0) {
  kable(unmatched_treatments)
} else {
  kable(tibble(status = "All non-control treatments are matched."))
}
```

## Molecular Profiles

The molecular profiles are stored as `ExpressionSet` objects in the `XevaSet`.
The matrix dimensions below are the feature-by-sample dimensions for each
profile.

```{r molecular-profile-dimensions}
profile_summary <- lapply(
  names(mcgill@molecularProfiles),
  function(profile_name) {
    profile <- mcgill@molecularProfiles[[profile_name]]
    feature_data <- Biobase::fData(profile)
    data.frame(
      profile = profile_name,
      features = nrow(exprs(profile)),
      samples = ncol(exprs(profile)),
      feature_id_example = rownames(exprs(profile))[[1]],
      feature_metadata = paste(colnames(feature_data), collapse = ", "),
      stringsAsFactors = FALSE
    )
  }
) |>
  bind_rows()

kable(profile_summary)
```

Gene-level molecular profiles are annotated with metadata from the current
GENCODE release used by the pipeline. RNAseq retains the original Ensembl
feature identifier, mutation and copy-number profiles retain their source
gene-symbol feature labels, and fusion profiles retain fusion-pair features
with separate GENCODE annotation for the left and right genes. Segment profiles
remain genomic interval features.

```{r molecular-feature-annotation}
gene_feature_summary <- lapply(
  names(mcgill@molecularProfiles),
  function(profile_name) {
    profile <- mcgill@molecularProfiles[[profile_name]]
    feature_data <- Biobase::fData(profile) |>
      as_tibble()

    if ("gencode.mapped" %in% colnames(feature_data)) {
      mapped <- feature_data$gencode.mapped
      if (!is.logical(mapped)) {
        mapped <- as.character(mapped) == "TRUE"
      }

      return(tibble(
        profile = profile_name,
        feature_set = "feature",
        features = nrow(feature_data),
        mapped_features = sum(mapped, na.rm = TRUE),
        unmapped_features = sum(!mapped, na.rm = TRUE),
        mapped_fraction = mapped_features / features,
        gencode_release = paste(
          sort(unique(na.omit(feature_data$gencode.release))),
          collapse = ", "
        ),
        example_gencode_symbol = dplyr::first(
          na.omit(feature_data$gencode.symbol)
        )
      ))
    }

    if (
      all(c(
        "gencode.left.mapped",
        "gencode.right.mapped"
      ) %in% colnames(feature_data))
    ) {
      bind_rows(
        lapply(c("left", "right"), function(side) {
          mapped <- feature_data[[paste0("gencode.", side, ".mapped")]]
          if (!is.logical(mapped)) {
            mapped <- as.character(mapped) == "TRUE"
          }

          tibble(
            profile = profile_name,
            feature_set = paste0(side, "_gene"),
            features = nrow(feature_data),
            mapped_features = sum(mapped, na.rm = TRUE),
            unmapped_features = sum(!mapped, na.rm = TRUE),
            mapped_fraction = mapped_features / features,
            gencode_release = paste(
              sort(unique(na.omit(feature_data[[paste0(
                "gencode.",
                side,
                ".release"
              )]]))),
              collapse = ", "
            ),
            example_gencode_symbol = dplyr::first(
              na.omit(feature_data[[paste0("gencode.", side, ".symbol")]])
            )
          )
        })
      )
    } else {
      NULL
    }
  }
) |>
  bind_rows()

kable(gene_feature_summary)
```

```{r unmapped-genes}
unmapped_genes_path <- data_path("results", "unmapped_genes.csv")
if (file.exists(unmapped_genes_path)) {
  unmapped_genes <- fread(unmapped_genes_path)
  if (nrow(unmapped_genes) > 0) {
    unmapped_genes |>
      count(profile, feature_column, name = "unmapped_features") |>
      arrange(desc(unmapped_features)) |>
      kable()
  } else {
    kable(tibble(status = "All gene-level molecular features are mapped."))
  }
} else {
  kable(tibble(status = "Unmapped gene report has not been generated yet."))
}
```

The model-to-molecular map links model records to patient or PDX molecular
samples. A model can map to both patient and PDX molecular samples for the same
profile type.

```{r molecular-map-coverage}
map_by_datatype <- mod_map |>
  distinct(model.id, biobase.id, source, mDataType) |>
  count(mDataType, source, name = "mappings") |>
  arrange(mDataType, source)

kable(map_by_datatype)
```

```{r molecular-model-coverage}
model_coverage <- mod_map |>
  distinct(model.id, mDataType) |>
  count(mDataType, name = "mapped_model_records") |>
  arrange(mDataType)

case_coverage <- model_df |>
  transmute(
    patient.id,
    RNAseq.PDX = !is.na(RNAseq.PDX),
    WGS.PDX = !is.na(WGS.PDX)
  ) |>
  distinct() |>
  summarise(
    patients = n_distinct(patient.id),
    patients_with_rnaseq_pdx = sum(RNAseq.PDX, na.rm = TRUE),
    patients_with_wgs_pdx = sum(WGS.PDX, na.rm = TRUE)
  )

kable(model_coverage)
kable(case_coverage)
```

## Response Summaries

The pipeline computes model-level and batch-level response metrics using the
`Xeva` package.

```{r model-response-availability}
model_metric_summary <- tibble(
  metric = colnames(model_metrics)[colnames(model_metrics) != "model.id"],
  available_models = colSums(!is.na(model_metrics[
    colnames(model_metrics) != "model.id"
  ])),
  pct_models = round(100 * colMeans(!is.na(model_metrics[
    colnames(model_metrics) != "model.id"
  ])), 1)
)

kable(model_metric_summary)
```

```{r batch-response-availability}
batch_metric_summary <- tibble(
  metric = colnames(batch_metrics)[colnames(batch_metrics) != "batch.name"],
  available_batches = colSums(!is.na(batch_metrics[
    colnames(batch_metrics) != "batch.name"
  ])),
  pct_batches = round(100 * colMeans(!is.na(batch_metrics[
    colnames(batch_metrics) != "batch.name"
  ])), 1)
)

kable(batch_metric_summary)
```

### Response Metric Distributions

```{r response-distributions, fig.height=6}
auc_df <- model_metrics |>
  select(AUC) |>
  filter(!is.na(AUC), AUC > 0)

slope_df <- model_metrics |>
  select(slope) |>
  filter(!is.na(slope), between(slope, -90, 90))

bar_df <- model_metrics |>
  select(best.average.response) |>
  filter(!is.na(best.average.response))

base_hist_theme <- theme_minimal(base_size = 12) +
  theme(
    panel.grid.minor = element_blank(),
    plot.title = element_text(face = "bold", size = 12),
    axis.title.x = element_text(size = 11),
    axis.title.y = element_text(size = 11)
  )

p_auc <- ggplot(auc_df, aes(x = AUC)) +
  geom_histogram(bins = 30, fill = "#3b82b8", color = "white") +
  scale_x_continuous(labels = label_number(big.mark = ",")) +
  labs(title = "AUC", x = "AUC", y = "Models") +
  base_hist_theme

p_slope <- ggplot(slope_df, aes(x = slope)) +
  geom_histogram(bins = 30, fill = "#3b82b8", color = "white") +
  labs(title = "Slope", x = "Angle (degrees)", y = "Models") +
  base_hist_theme

p_bar <- ggplot(bar_df, aes(x = best.average.response)) +
  geom_histogram(bins = 30, fill = "#3b82b8", color = "white") +
  labs(
    title = "Best Average Response",
    x = "Percent change from baseline",
    y = "Models"
  ) +
  base_hist_theme

(p_auc + p_slope + p_bar) +
  plot_layout(ncol = 3)
```

`AUC`, `slope`, and `best.average.response` summarize different aspects of
tumor growth. Lower `best.average.response` values indicate stronger inhibition
or shrinkage relative to baseline.

## mRECIST Summary

The `mRECIST` matrix summarizes response by treatment and patient.

```{r mrecist-table}
mrecist_matrix <- summarizeResponse(
  mcgill,
  response.measure = "mRECIST",
  group.by = "patient.id",
  tissue = "Breast Cancer"
)

dim(mrecist_matrix)
```

```{r mrecist-preview}
kable(
  mrecist_matrix[seq_len(min(8, nrow(mrecist_matrix))),
    seq_len(min(8, ncol(mrecist_matrix))),
    drop = FALSE
  ]
)
```

```{r mrecist-plot, fig.width=11, fig.height=6}
plotmRECIST(
  mrecist_matrix,
  control.name = "untreated",
  name = "Breast cancer",
  sort = FALSE
)
```

## Example Growth Curves

The batch-level plot pairs a treatment arm with its control arm for the same
patient.

```{r batch-example}
sample_batch <- batch_ids[grepl("Cisplatin", batch_ids)][[1]]
batch_info <- batchInfo(mcgill, batch = sample_batch)[[1]]

kable(tibble(
  field = c("Batch", "Control arm", "Treatment arm"),
  value = c(batch_info$batch.name, batch_info$control, batch_info$treatment)
))
```

```{r plot-batch-growth, fig.height=5}
plotPDX(
  mcgill,
  batch = sample_batch,
  vol.normal = TRUE,
  max.time = 30
)
```

## Waterfall Plot

Waterfall plots rank models from strongest response to weakest response for a
given treatment.

```{r waterfall-drug}
waterfall_drug <- model_df |>
  filter(drug != "untreated") |>
  count(drug, sort = TRUE) |>
  slice(1) |>
  pull(drug)

waterfall_drug
```

```{r waterfall-plot, fig.width=10, fig.height=5}
waterfall(
  mcgill,
  drug = waterfall_drug,
  res.measure = "best.average.response"
)
```

## Session Info

```{r session-info}
sessionInfo()
```
